Solving Graph Homomorphism and Subgraph Isomorphism Problems Faster Through Clique Neighbourhood Constraints

### 2021-01-19 paper/paper.tex
Last tweaks

The current state of the art is the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020}, which
is a dedicated constraint programming solver for subgraph-finding problems. Much of its performance
comes from inference strategies based upon degrees and neighbourhood degree sequences
\cite{DBLP:journals/constraints/ZampelliDS10}, counting paths between vertices
\cite{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}, and cardinality reasoning
\cite{DBLP:journals/ai/Solnon10}, which can be used to eliminate many infeasible candidate
assignments without search. Like most other solvers, it can handle both the non-induced version of
the problem, as well as the traditional induced variant, and users can select whichever variant
better fits their application. It also supports directed edges, and can use an external constraint
solver to handle other extensions to the basic notion of a graph, such as temporal networks and
multi-graphs.

@@

\paragraph{Loops.} A vertex which is adjacent to itself is called a \emph{loop}. By a careful reading
of the definition, any homomorphism must map loops onto loops---and indeed, the Glasgow Subgraph
Solver and this paper take this approach (although some other algorithm implementations do not). It
therefore follows that for the homomorphism \emph{decision} problem where there are no injectivity
constraints, any problem instance which has a loop in the codomain graph is trivially satisfiable,
although the \emph{counting} problem remains \#\P-hard \cite{DBLP:journals/rsa/DyerG00}.

@@

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two (see \cref{figure:counterexamples}, top right). While standard injectivity
prevents any kind of merging, local injectivity prevents two vertices from being merged if they are
both in the neighbourhood of the same vertex.  So the second property does not hold in full
generality, because, for example, a path of length three may be mapped onto a triangle whilst
respecting local injectivity (\cref{figure:counterexamples}, bottom left).  However, we observe that
a weaker result holds:

@@

Rather than using distance properties directly, \citet{DBLP:conf/cp/McCreeshP15} introduced
the notion of \emph{supplemental} graphs, with the idea that a constraint programming algorithm can
search for a mapping which is simultaneously a subgraph isomorphism between several different pairs
of graphs. Let $G^d$ be the graph with the same set of vertices as $G$, but an edge between vertices
$v$ and $w$ if the distance between $v$ and $w$ in $G$ is at most $d$.  Similarly, let $G^{n,\ell}$
be the graph with the same set of vertices as $G$, but an edge between vertices $v$ and $w$ if there
are at least $n$ simple paths of length exactly $\ell$ between $v$ and $w$ in $G$. Following on from
\cref{proposition:distances,proposition:paths}, we generalise the result of
\citeauthor{DBLP:conf/cp/McCreeshP15} as follows.

@@

\paragraph{Cliques and homomorphisms.} Finally, we come to the main result of this paper. Observe
that until now, we have not seen any unary constraints which are valid for the general homomorphism
problem. All existing constraint programming approaches for the subgraph isomorphism problem begin
by branching on the variable which has the smallest domain---typically this will be a domain for a
domain vertex of high degree, or which has many high degree neighbours. Furthermore, once one such
guessed assignment has been made, adjacency propagation means many other domains will be
substantially reduced in size, making subsequent branching choices simpler.  However, for
homomorphism problems, if we cannot find any implied unary constraints then every domain will
initially be of the same size, which will make it much harder for a solver to know where to start.
This motivates the following observation.

@@

\paragraph{Loops.} For the homomorphism decision problem, we modified the solver so that if the
codomain graph contains a loop, we immediately return a satisfying assignment mapping every vertex
to the loop.

@@

Implementing clique constraints required substantially more work. The Glasgow Subgraph Solver
contains a maximum clique implementation, which is based upon a variation of
\citet{DBLP:conf/walcom/TomitaSHTW10}'s MCS algorithm which
\citet{DBLP:journals/algorithms/Prosser12} calls ``MCSa1'', with bit-parallelism and an altered
branching scheme for faster optimality proofs \cite{DBLP:conf/cp/McCreeshP14}. The implementation
also makes use of a shortcut due to \citet{DBLP:journals/jco/BatsynGMP14}, which allows certain
cliques to be detected without branching---this turns out to be particularly useful in this setting,
making many of the clique instances generated solvable using only a single greedy colouring step.
Initially we used this solver to perform domain filtering as a preprocessing step. Preliminary
experiments showed that a na{\"\i}ve approach, which simply calculates the maximum clique size for
the neighbourhood of each domain vertex and each codomain vertex, would add as much as three minutes
of preprocessing time to some problem instances which could otherwise be solved in a few seconds. We
therefore use a more intelligent approach, which we now describe.

Assuming a problem instance is not detected as obviously unsatisfiable, we calculate the maximum
clique size for the neighbourhood of every domain vertex in turn. However, having found the maximum
clique with $k$ vertices in the neighbourhood of a vertex $p$, we remember for every other vertex in
the maximum clique that its neighbourhood maximum clique size is at least $k$. This can be used to
accelerate subsequent clique solver calls, by starting the branch and bound algorithm with an
initial incumbent size of $k$ rather than zero.

@@

Given the complexity of implementations of modern solvers, a critical question is how we can be sure
that they are producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and
clique algorithms are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they
can output a mathematical proof that they have reached a correct answer by sound reasoning, and this
proof log can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}.

@@

In the top left plot of \cref{figure:plots} we show the cumulative number of instances solved over
time, for each of the three problem variants, using the Glasgow Subgraph Solver with and without our
clique-filtering added. For the homomorphism problem, we also compare with distance filtering
enabled or disabled. For subgraph isomorphism, we also compare against the PathLAD
\cite{DBLP:conf/lion/KotthoffMS16}, VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
\cite{DBLP:journals/bmcbi/BonniciGPSF13} solvers. Because every Glasgow Subgraph Solver result is so
much stronger than the other three solvers, it is hard to get any further information from this
plot. We therefore re-plot zoomed in results on the second row of \cref{figure:plots}, focusing each
problem variant in turn. The results show that for the homomorphism problem, using either clique or
distance filtering gives a clear improvement to performance, and using both together is better
still. For locally injective homomorphisms and subgraph isomorphism, clique filtering also gives an
improvement.

To quantify these gains, the top centre plot of \cref{figure:plots} shows the
additional number of instances solved at any given time (that is, the vertical distance between
two cumulative plots) when enabling additional filtering. We see that for subgraph isomorphism, at
the one hour timeout we solve 22 additional instances using clique filtering; for locally injective
homomorphisms, 21 additional instances; and for homomorphisms, 114, 146 or 193 additional instances
using just distance filtering, just clique filtering, or both. Alternatively, we can measure the
horizontal distance between plots for a given choice of timeout $t$ as follows. Let $t'$
be the last time no greater than $t$ where the slower algorithm solved an instance, and let $u$ be
the lowest amount of time needed for the faster algorithm to solve the same number of instances.
Then the aggregate speedup, which we plot in the top right of \cref{figure:plots}, is the ratio of
$t'$ to $u$ or $u$ to $t'$ as appropriate. At the timeout, we see aggregate speedups of 2.2 for
locally injective homomorphisms, 3.2 for subgraph isomorphism, and 47.5, 131.7, and 803.5 for
homomorphisms.

The crossover point, where clique filtering begins to pay off, is at around the 100ms mark for all
homomorphism variants, and at around one second for subgraph isomorphism. This does not mean that
this much time was spent in clique-finding, but rather that this is how long it takes for an
algorithm with increased filtering to catch up from a slower start. In fact, for domain graphs, no
maximum clique call took even one millisecond to complete (and none required more than two hundred
recursive calls from the clique solver), and more time was spent setting up the necessary data
structures to represent the graphs than on the solving itself. For codomain graphs, 20,838,008 of
the clique problems we solved took under a millisecond, and the remaining 279 calls took between one
and four milliseconds, with no instance requiring more than a thousand recursive calls to solve.
Although it may seem counter-intuitive to try to speed up the solving of a hard problem by first
solving many other hard problems, these values demonstrate that clique-solving on these smaller
graphs is \emph{much} easier in practice than solving the larger problem, and so is worthwhile.

Finally, \cref{figure:scatters}, plots instance-by-instance comparisons of the runtimes. The
point styles in these plots show different families of benchmark instance, and demonstrate
that the benefits of filtering are not restricted to one application.

@@

These results show that constraint programming approaches are viable for a wider range of subgraph
mapping finding problems than had previously been considered, which will increase the versatility of
subgraph solvers for application users going forwards. We also saw that using clique-finding as a
filtering step could give moderate to very large speedups for different problem variants---although
only through careful engineering choices, to mitigate the potential cost of solving \NP-hard
problems. We would be interested to see if filtering based upon other hard problems could give
similar benefits---for example, homomorphisms also interact in potentially helpful ways with various
colouring properties.

>>>>

The current state of the art is the Glasgow Subgraph Solver \cite{DBLP:conf/gg/McCreeshP020}, which
is a dedicated constraint programming solver for subgraph-finding problems. Much of its performance
comes from inference strategies based upon degrees and neighbourhood degree sequences
\cite{DBLP:journals/constraints/ZampelliDS10}, counting paths between vertices
\cite{DBLP:conf/cp/AudemardLMGP14,DBLP:conf/cp/McCreeshP15}, and cardinality reasoning
\cite{DBLP:journals/ai/Solnon10}, which can be used to eliminate many infeasible candidate
assignments without search. Like most other solvers, it can handle both the non-induced version of
the problem and the traditional induced variant, and users can select whichever variant
better fits their application. It also supports directed edges, and can use an external constraint
solver to handle other extensions to the basic notion of a graph, such as temporal networks and
multi-graphs.

@@

\paragraph{Loops.} A vertex which is adjacent to itself is called a \emph{loop}. By a careful reading
of the definition, any homomorphism must map loops onto loops---and indeed, the Glasgow Subgraph
Solver and this paper take this approach (although some other algorithm implementations do not). It
therefore follows that for the homomorphism \emph{decision} problem where there are no injectivity
constraints, any problem instance which has a loop in the codomain graph is trivially satisfiable
(although the \emph{counting} problem remains \#\P-hard \cite{DBLP:journals/rsa/DyerG00}).

@@

It is easy to verify that the first of these two properties is valid for \emph{any} homomorphism.
The second property does not hold for homomorphisms in general: for example, a pair of vertices
connected by three paths of length two may be mapped onto a pair of vertices connected by a single
path of length two (see \cref{figure:counterexamples}, top right). The second property also does not
hold in full generality for local injectivity, because, for example, a path of length three may be
mapped onto a triangle (\cref{figure:counterexamples}, bottom left): while standard injectivity
prevents any kind of merging, local injectivity only prevents two vertices from being merged if they
are both in the neighbourhood of the same vertex. However, we observe that a weaker result does
hold:

@@

Rather than using distance properties directly as \citet{DBLP:conf/cp/AudemardLMGP14} did,
\citet{DBLP:conf/cp/McCreeshP15} introduced the notion of \emph{supplemental} graphs, with the idea
that a constraint programming algorithm can search for a mapping which is simultaneously a subgraph
isomorphism between several different pairs of graphs. Let $G^d$ be the graph with the same set of
vertices as $G$, but an edge between vertices $v$ and $w$ if the distance between $v$ and $w$ in $G$
is at most $d$.  Similarly, let $G^{n,\ell}$ be the graph with the same set of vertices as $G$, but
an edge between vertices $v$ and $w$ if there are at least $n$ simple paths of length exactly $\ell$
between $v$ and $w$ in $G$. Following on from \cref{proposition:distances,proposition:paths}, we
generalise the result of \citeauthor{DBLP:conf/cp/McCreeshP15} as follows.

@@

\paragraph{Cliques and homomorphisms.} We now discuss the main new technique of this paper. Observe
that so far, we have not seen any unary constraints which are valid for the general homomorphism
problem. All existing constraint programming approaches for the subgraph isomorphism problem begin
by branching on the variable which has the smallest domain---typically this will be a domain for a
vertex of high degree, or which has many high degree neighbours. Furthermore, once one such
guessed assignment has been made, adjacency propagation means many other domains will be
substantially reduced in size, making subsequent branching choices simpler.  However, for
homomorphism problems, if we cannot find any implied unary constraints then every domain will
initially be of the same size, which will make it much harder for a solver to know where to start.
This motivates the following.

@@

\paragraph{Loops.} For the homomorphism decision problem, we modified the solver so that if the
codomain graph contains a loop, we immediately return a satisfying assignment mapping every vertex
to one of these loops.

@@

Implementing clique constraints required substantially more work. The Glasgow Subgraph Solver
contains a maximum clique implementation, which is based upon a variation of
\citet{DBLP:conf/walcom/TomitaSHTW10}'s MCS algorithm which
\citet{DBLP:journals/algorithms/Prosser12} calls ``MCSa1'', with bit-parallelism
\cite{DBLP:journals/ol/SegundoMRH13} and an altered
branching scheme for faster optimality proofs \cite{DBLP:conf/cp/McCreeshP14}. The implementation
also makes use of a shortcut due to \citet{DBLP:journals/jco/BatsynGMP14}, which allows certain
cliques to be detected without branching---this turns out to be particularly useful in this setting,
making many of the clique instances generated solvable without branching.
Initially we used this solver to perform domain filtering as a preprocessing step. Preliminary
experiments showed that a na{\"\i}ve approach, which calculates the maximum clique size for
the neighbourhood of each domain vertex and each codomain vertex, would add as much as three minutes
of preprocessing time to some problem instances which could otherwise be solved in a few seconds. We
therefore invested more engineering effort, as follows.

Assuming a problem instance is not detected as obviously unsatisfiable, we calculate the maximum
clique size for the neighbourhood of every domain vertex in turn. However, having found a maximum
clique with $k$ vertices in the neighbourhood of a vertex $p$, we remember for every other vertex in
this maximum clique that its neighbourhood maximum clique size is at least $k$. This can be used to
accelerate subsequent clique solver calls, by starting the branch and bound algorithm with an
initial incumbent size of $k$ rather than zero.

@@

Given the complexity of modern solvers, a critical question is how we can be sure
that they are producing correct answers. The Glasgow Subgraph Solver's subgraph isomorphism and
clique algorithms are both \emph{certifying} \cite{DBLP:journals/csr/McConnellMNS11}: that is, they
can output a mathematical proof that they have reached a correct answer by sound reasoning, and this
proof log can be verified by the VeriPB proof checking tool
\cite{DBLP:conf/cp/GochtMMNPT20,DBLP:conf/ijcai/GochtMN20}.

@@

In the top left plot of \cref{figure:plots} we show the cumulative number of instances solved over
time, for each of the three problem variants, using the Glasgow Subgraph Solver with and without our
clique-filtering added. For the homomorphism problem, we also compare with distance filtering
enabled or disabled. For subgraph isomorphism, we further compare against the PathLAD
\cite{DBLP:conf/lion/KotthoffMS16}, VF2 \cite{DBLP:journals/pami/CordellaFSV04} and RI
\cite{DBLP:journals/bmcbi/BonniciGPSF13} solvers. To make the details of this plot readable, the
second row of \cref{figure:plots} zooms in on the results for each
problem variant in turn. We see that for the homomorphism problem, using either clique or
distance filtering gives a clear improvement to performance, and using both together is better
still. For both injective problems, clique filtering gives a more moderate improvement.

To quantify these gains, the top centre plot of \cref{figure:plots} shows the
additional number of instances solved at any given time (that is, the vertical distance between
two cumulative plots) when enabling additional filtering. We see that for subgraph isomorphism, at
the one hour timeout we solve 22 additional instances using clique filtering; for locally injective
homomorphisms, 21 additional instances; and for homomorphisms, 114, 146 or 193 additional instances
using just distance filtering, just clique filtering, or both. Alternatively, we can measure the
horizontal distance between plots for a given choice of timeout $t$ as follows. Let $t'$
be the last time no greater than $t$ where the slower algorithm solved an instance, and let $u$ be
the lowest amount of time needed for the faster algorithm to solve the same number of instances.
The \emph{aggregate speedup}, which we plot in the top right of \cref{figure:plots}, is the ratio of
$t'$ to $u$ or $u$ to $t'$ as appropriate. At the one hour timeout, we see aggregate speedups of 2.2
for locally injective homomorphisms, 3.2 for subgraph isomorphism, and 47.5, 131.7, and 803.5 for
homomorphisms.

The crossover point, where clique filtering begins to pay off, is at around the 100ms mark without
injectivity, and at around one second for the injective problems. This does not show how much time
was spent in clique-finding, but rather how long it takes for an algorithm with increased
filtering to catch up from a slower start. In fact, for domain graphs, no maximum clique call took
even one millisecond to complete (and none required more than two hundred recursive calls from the
clique solver), and more time was spent setting up the data structures to represent the
neighbourhood graphs than on the actual solving. For codomain graphs, 20,838,008 of the clique
problems we solved took under a millisecond, and the remaining 279 calls took between one and four
milliseconds, with no instance requiring more than a thousand recursive calls.  Although it may seem
counter-intuitive to try to speed up the solving of a hard problem by first solving many other hard
problems, these values demonstrate that clique-solving on these smaller graphs is \emph{much} easier
in practice than solving the larger problem, and so is worthwhile.

Finally, \cref{figure:scatters} plots instance-by-instance comparisons of the runtimes. The
point styles in these plots show different families of benchmark instance, and demonstrate
that the benefits of filtering are not restricted to one application. They also show that clique
filtering very rarely makes performance a lot worse, but often makes it a lot better.

@@

Our results have shown that constraint programming approaches are viable for a wider range of subgraph
mapping finding problems than had previously been considered, and will increase the versatility of
subgraph solvers for application users going forwards. We saw that using clique-finding as a
filtering step could give moderate to very large speedups for different problem variants---although
only through careful engineering choices, to mitigate the potential cost of solving \NP-hard
problems. We would be interested to see if filtering based upon other hard problems could give
similar benefits---for example, homomorphisms also interact in potentially helpful ways with various
colouring properties.
